Smart Drill-Down

### 2014-11-30 TableSummarization/Paper/sampling.tex
sampling reorg

\subsection{Dynamic Sampling for Large Tables}\label{sec:sampling}
We now describe our sampling schemes for improving the running time of our algorithm on tables that are too large to fit in main memory. First in Section~\ref{sec:sample-using}, we motivate the need for sampling. Then we describe a component of our system, called the {\em SampleHandler}, which is responsible for creating and maintaining samples of the table in memory, subject to user specified memory constraints. The SampleHandler maintains multiple samples corresponding to different parts of the table, which can be used depending on which rule the user decides to expand next. Then in Section~\ref{sec:sampling_algorithms}, we describe ways to efficiently allocate memory to different samples, so as to maximize the probability that we can respond to the next user operation without accessing the hard disk. 
\papertext{In the technical report,}
\techreporttext{Finally,}
we mention some additional optimizations we can make, and
describe how we can set the minimum sample size required from the SampleHandler.
\agp{revisit}

\subsubsection{Using samples to respond to user operations}\label{sec:sample-using}
Our greedy algorithm needs to make multiple passes over the entire table in order to find counts of rules. These passes can be very expensive if the table is large, especially if the table does not fit in main memory. If we want exact counts for rules, we have no choice but to read the entire table.

@@

Thus, we use sampling to trade-off a small amount of accuracy for a faster response time. Our system includes a {\em SampleHandler}, which is given a certain memory capacity $M$, and a minimum sample size $minSS$ (both specified by the user).
The $minSS$ parameter is the minimum number of sample tuples that are used to determine counts while running our greedy algorithm, and this parameter determines how accurate our count estimates will be. We later provide a way to find reasonable values of $minSS$.

% Example from situation in a previous table. And good and bad samples to keep there.
At all points, the SampleHandler maintains a set of samples in memory. For instance, it may keep a sample of tuples used to expand the first (trivial) rule, and another sample used to expand the rule last clicked on by the user. Each sample $s$ is an object with three attributes: A `filter' rule $f_s$, a scaling factor $N_s$ and a set $T_s$ of tuples from the table. The set $T_s$ consists of a $\frac{1}{N_s}$ uniformly sampled fraction of tuples covered by $f_s$. The scaling factor $N_s$ is used to translate the count of a rule on the sample into an estimate of the count over the entire table. The sum of $|T_s|$ over all samples $s$ is not allowed to exceed capacity $M$ at any point. 
% TODO: Give example of what samples we'd keep.

Whenever the user drills down on a rule $r$, our system calls the SampleHandler with argument $r$, which returns a sample $s$ whose filter value is given by $f_s = r$ and has $|T_s| \geq minSS$. Thus, the $T_s$ of the returned sample consists of a uniformly random set of tuples covered by $r$. The SampleHandler also computes $N_s$ when a sample is created. Then we run Algorithm~\ref{algo:best-rule-set} on sample $s$ (with a modified weight function in case the user clicked on a $\star$) to obtain the list of rules to display. The counts of the rules on the sample are multiplied by $N_s$ before being displayed, to get estimated counts on the entire table. In addition since the sample is uniformly random, we can also compute confidence intervals on the estimated count of each displayed rule, although we do not currently display the confidence intervals.

When the SamplerHandler gets called with argument $r$, it needs to find or create a sample with $r$ as the filter rule. At the beginning when it gets called with the empty rule as an argument, there are no samples in memory and it must make a pass through the data to generate a sample. Creating a new sample by making a pass through the table is called \textbf{Create} (further described below). At later stages, when there are potentially multiple samples available, there are multiple ways it could return a sample for rule $r$:
\begin{enumerate}
\item \textbf{Find:} If the SampleHandler finds an existing sample $s$ in memory, which has $r$ as its filter rule (i.e. $f_s = r$) and at least $minSS$ tuples ($|T_s| \geq minSS$, then it simply returns sample $s$. Algorithm~\ref{algo:best-rule-set} can then be run on $s$.

\item \textbf{Combine:} If \textbf{Find} doesn't work i.e. if the SampleHandler cannot find an existing sample with filter $r$ and $\geq minSS$ tuples, then it looks at all existing samples $s^{\prime}$ such that $f_{s^{\prime}}$ is a sub-rule of $r$. If the set of all tuples that are covered by $r$, from all such $T_{s^{\prime}}$'s combined, exceeds $minSS$ in size, then we can simply treat that set as our sample for rule $r$.
We can show that tuples that are covered by $r$, from the combination of $T_{s^{\prime}}$s, follow a uniform distribution. That is, each table tuple $t$ that is covered by $r$ is equally likely to appear in a $T_{s^{\prime}}$.

In addition, the SampleHandler needs to return a sample of size equal to $minSS$. It can create a new sample of $< minSS$ size if there are additional tuples covered by $r$ present in existing samples, or $> minSS$ (if enough memory is available). Making a larger sample is advantageous not only to get higher accuracy, but also because, when the user later drills down on a sub-rule $r^{\prime}$ of $r$, having a large $r$ sample increases the chance that the \textbf{Combine} strategy will work for $r^{\prime}$, which can let us avoid making another expensive pass through the table. For example, if $minSS = 500$, but we get a size $2000$ sample $s$ for the empty rule, then when the user clicks on one of its sub-rules, say $r$, there is a good chance the $2000$ tuples from $T_s$ contain at least $500$ tuples covered by $r$ and that allows us to display the rule-list expanding $r$ quickly instead of making another pass through the table. We describe our algorithms for deciding what sized samples to create, and what samples to shrink, in Section~\ref{sec:sampling_algorithms}.

\item \textbf{Delete/Shrink:} When the SampleHandler uses \textbf{Create}, it may not have enough memory left to create a new sample. In that case, the SampleHandler needs to delete or shrink an existing sample. It could use a heuristic, like a LRU (Least-Recently-Used) policy to decide which sample to delete or shrink. Note that shrinking a sample by randomly deleting some tuples (even if the sample size goes below $minSS$) may be better than deleting it entirely, because in future, we may be able to recreate a sample using \textbf{Combine} on the existing shrinked sample along with other samples.
\end{enumerate}

We describe our algorithm for deciding what samples to shrink, and what sized samples to create, in Section~\ref{sec:sampling_algorithms}.

\textbf{Pre-fetching:} When the user clicks on rule $r$ (on the rule itself or on a $\star$ in the rule), we need to get a sample, run the Algorithm~\ref{algo:best-rule-set}, and display a rule-list to the user. If we use \textbf{Find} or \textbf{Combine}, then we can display the rule-list much faster because we don't have to read the entire table. But after expanding $r$, there is a high chance that the user goes further and drills down on one of the sub-rules $r^{\prime}$ of $r$. We may not be able to use \textbf{Find} or \textbf{Combine} on $r^{\prime}$ with the existing samples. So while the user is reading the current rule-list obtained from drilling down on $r$, we can start making a pass through the table to create a bigger sample of $r$ in the background. That way, when the user expands $r^{\prime}$, some of the newly loaded tuples for $r$ will also be covered by $r^{\prime}$, increasing the chance that we can use \textbf{Combine} on $r^{\prime}$, and reduce our response time.
In addition, while we are making the pass in the background, we can find the exact counts for currently displayed rules (which only have estimated counts shown), and update them when our pass is complete.

\subsubsection{Algorithms for deciding what to sample}\label{sec:sampling_algorithms}
We now discuss algorithms for the SampleHandler to decide what sizes of Samples to maintain in memory. Briefly, we want to determine sample sizes that maximize the probability that the next user click can be answered using the existing samples, without reading from the hard disk. We leverage techniques from approximation algorithms and optimization theory.

At any stage, we have a tree $U$ displayed to the user, with each node of the tree corresponding to a displayed rule. The tree is formed as follows: The root of the tree corresponds to the trivial rule. And suppose the user expands a node with rule $r$, resulting in rules $r_1, r_2, .. r_k$ being displayed. Then we add children nodes to the expanded node, corresponding to rules $r_1, r_2, .. r_k$. Note that multiple nodes of the tree may correspond to the same rule. Thus, even though the structure displayed to the user is a tree, the set of rules displayed forms a partially ordered set (poset) using the sub-rule ordering. Internal nodes of $U$ are ones that have been expanded (drilled down on), while leaves are nodes that have not been expanded. Let $L$ be the set of leaves. Each leaf is something that the user can potentially expand in the next step, and thus we would like to have pre-fetched samples for rules corresponding to leaf nodes.
In the rest of this section, we abuse notation to use the term rule to refer to any node that corresponds to that rule.

We assume that we have a probability distribution over leaves, which assigns a probability that each leaf will be the next one to be expanded. In the absence of additional data, we can assume a uniform probability distribution. That is, we can assume that every leaf is equally likely to be expanded next. If we have data on past user behaviour available, we can use Machine Learning on node features such as `node depth in tree', `weight of node rule' and `distance from last expanded node' to get a better probability estimate of each node being expanded next.

When the SamplerHandler uses \textbf{Create} for a rule $r$, it needs to make a pass through the entire table, as well as potentially shrink some existing samples to free up memory. Since making a pass over the the table from the hard disk is usually a bottleneck, it can also do things like creating samples for rules other than $r$, and augmenting existing samples, in the same pass. Hence, we assume that in a \textbf{Create} phase, the SampleHandler not only creates one new sample for $r$, but also potentially creates other new samples, or resizes existing samples. For each displayed rule $r^{\prime}$, it may create a new sample for $r^{\prime}$ in the same pass through the table. Based on the current leaves and their expansion probabilities, it finds an optimal set of sizes of samples to create for each displayed rule. More specifically, for every displayed rule $r^{\prime}$, it determines an integer $n_{r^{\prime}}$, and then creates a fresh sample $s_{r^{\prime}}$ with $f_{s_{r^{\prime}}} = r^{\prime}$ and $|T_{s_{r^{\prime}}}| = n_{r^{\prime}}$ while making its pass through the table.

We choose the $n_{r^{\prime}}$ values so as to maximize the probability that the next user drill down can be satisfied using samples available in memory. We define some terms required to formalise this problem next.

Let the `selectivity' of a rule be the fraction of tuples in $T$ that are covered by the rule. For each pair of rules $r_1, r_2 \in U$ such that $r_1$ is a sub-rule of $r_2$, we can estimate the ratio of selectivities of $r_1$ and $r_2$ using existing samples. Call this quantity $S(r_1, r_2)$. We define $S(r_1, r_2)$ to be $0$ if $r_1$ is not a sub-rule of $r_2$. If the same rule occurs in multiple nodes $r_1$, $r_2$ of the tree, then $S(r_1, r_2)$ is naturally $1$. Then if we have an $n_r$ sized sample with filter $r$ for each $r \in U$, the expected number of tuples covered by $r^{\prime} \in L$ is given by $$\sum_{r \in U} S(r, r^{\prime})n_r$$ For any $r^{\prime}$, define the function {\em ess} (for `effective sample size') as below: 
\begin{definition}\label{def:ess}
$$ess(r^{\prime}) = \sum_{r \in U} S(r, r^{\prime})n_r$$ 
\end{definition}
If $ess(r^{\prime}) \geq minSS$ then if the user expands $r^{\prime}$, we display the next rule list using \textbf{Find} or \textbf{Combine}, instead of having to make another pass through the table. We wish to maximize the probability that we can respond to the next user expansion without making another pass. We now formally define our problem below:
\begin{problem}\label{prob:sample-sizes}
Given a tree of rules $U$ with leaves $L$, a probability distribution $p$ over $L$, an integer $M$, and selectivity ratio $S(r_1, r_2)$ for each $r_1, r_2 \in U$, 
choose an integer $n_r \geq 0$ for each $r \in U$ so as to $\textrm{maximize}$ :
$$\sum_{r^{\prime} \in L} p_{r^{\prime}}I_{ess(r^{\prime}) \geq minSS}$$
where the $I$'s are indicator variables, subject to :
$$\sum_{r \in U} n_r < M$$
\end{problem}
Problem~\ref{prob:sample-sizes} is non-linear and non-convex because of the indicator variables. If the tree $U$ is not too big, then it might be feasible to use an exponential algorithm to solve the problem. We can show that Problem~\ref{prob:sample-sizes} is NP-Hard by reducing the knapsack problem to it. 
\begin{lemma}\label{lemma:sampling-hardness}
Problem~\ref{prob:sample-sizes} is NP-Hard.
\end{lemma}
\begin{proof}(Sketch)
Suppose we are given an instance of the knapsack problem with $m$ objects, with the $i^{th}$ object having weight $w_i$ and value $v_i$. We are also given a weight limit $W$, and our objective is to choose a set of objects that maximizes value and has total weight $< W$. We will reduce this to an instance of Problem~\ref{prob:sample-sizes}.

We first scale the $w_i$s and $W$ such that all $w_i$s are $< 1$, without effectively changing the problem. For Problem~\ref{prob:sample-sizes}, we set $M$ to $(m + W) \times minSS$. Tree $U$ has $m$ special nodes $r_1, r_2, ... r_m$, and each $r_i$ has two children $r_{i,1}, r_{i,2}$. These $2m$ children are all leaves, and all leaves other than these have expansion probability $0$. 
The $S$ values are such that $\forall i \in \left\lbrace 1, 2, ..m\right\rbrace, j \in \left\lbrace 1, 2 \right\rbrace : (S(x, r_{i,j}) \neq 0 \Rightarrow x = r_{i,j} || x = r_i)$. In reality, the $S$ values cannot be exactly zero when the first argument is the trivial rule, but we can make it small enough such that any optimal solution will set $n_r = 0$ when $r$ is not a special node or its child. Thus, each each leaf $r_{i,j}$ gets tuples either from its own sample, or the sample of its parent, and from nowhere else. Thus, $ess(r_{i,j}) = n_{r_{i,j}} + n_{r_i}S(r_i, r_{i,j}) \forall 1 \leq i \leq m, j \in \left\lbrace 1,2 \right\rbrace$. In addition, $S(r_i, r_{i,1}) = 1$ (again, it cannot be exactly $1$, but can be brought arbitrarily close to $1$), and $S(r_i, r_{i,2}) = 1 - w_i$. Finally, for each i, $p_{r_{i,1}} = \frac{2}{2m+1}$ and $p_{r_{i,2}} = \frac{v_i}{(2m+1)\sum_{j=1}^{m}v_i}$. Thus, each individual $p_{r_{i,1}}$ value is higher than all $p_{r_{i,2}}$ values combined, and $M$ is high enough to cover all $r_{i,1}$. Thus, in any optimal solution, $ess(r_{i,1})$ must be $= minSS$, and we're left to decide which $i$'s should also have $ess(r_{i,2}) = minSS$. For all $i$, we must have either $n_{r_i} = n_{r_{i,2}} = 0 \land n_{r_{i,1}} = minSS$ (iff $ess(r_{i,2}) < minSS$) or $n_{r_i} = minSS \land n_{r_{i,1}} = 0 \land n_{r_{i,2}} = minSS (1 - S(r_i, r_{i,2})) = minSS \times w_i$ (iff $ess(r_{i,2}) = minSS$). The latter option consumes $w_i \times minSS$ extra memory and gives an extra $\frac{v_i}{(2m+1)(\sum_{j=1}^{m}v_j)}$ probability value. Thus, having $ess(r_{i,2}) = minSS$ is equivalent to picking object $i$ from the knapsack problem (as it consumes additional memory $\propto w_i$ and gives additional probability value $\propto v_i$). Moreover the additional memory available (on top of the $m \times minSS$ required to cover all $r_{i,1}$s) is $W\times minSS$. Hence solving Problem~\ref{prob:sample-sizes} with the above $U$, $S$, $p$ and picking the set of $i$'s for which $ess(r_{i,2}) = minSS$ gives a solution to the instance of the knapsack problem. 
\end{proof}

However, with an additional simplifying assumption, we can make the problem approximately solvable. The assumption is: For each $r \in L$, we assume that it will get tuples only from samples obtained for itself (filter $= r$) and its immediate parent. That is, we set $S(r_1, r_2)$ to be zero if $r_1 \neq r_2$ and $r_2$ is not a child of $r_1$. This is similar to what we had for tree $U$ in our proof of Lemma~\ref{lemma:sampling-hardness}. $ess(r)$ now becomes $ess(r^{\prime}) = n_{r^{\prime}} + n_rS(r, r^{\prime})$ where $r$ is the parent of $r^{\prime}$.

Now consider a rule in $r_0 \in U \setminus L$ along with all its children. Let $M_{r_0}$ denote the set containing $r_0$ and all its leaf children. Then by our simplification, the number of tuples $n_{r^{\prime}}$ for any rule in $r^{\prime} \in M_{r_0}$ only affects the ess value of rules in $M_{r_0}$. This allows us to effectively split the problem into multiple subproblems,  with one subproblem per $M_{r_0}$. Thus, for each non-leaf rule $r_0$ and all its children, we compute all `locally optimal' assignments of $n_r \mid r \in M_{r_0}$. Locally optimal means that we cannot get a higher value of `probability value' $\sum_{r \in M_{r_0}} p_rI_{ess(r) \geq minSS}$ for the same `sampling cost' $\sum_{r\in M_{r_0}} n_r$. Then we can use dynamic programming to combine the locally optimal solutions of different $M_{r_0}$s. We describe both these steps in detail below:

Let $r_0 \in U \setminus L$. Let $d$ be the number of leaf children of $r_0$. Let the children be $r_1, r_2, ... r_d$. For any child $r_i$, $n_{r_i}$ only contributes to it's own $ess$, whereas $n_{r_0}$ contributes to the $ess$ of all children $r_1, ... r_d$. Given a value of $n_{r_0}$, in a locally optimal solution, each child $r_i$ must satisfy:
\squishlist
\item If $n_{r_0}S(r_0, r_i) \geq minSS$, then $n_{r_i} = 0$ because otherwise, decreasing $n_{r_i}$ to $0$ would lower its sampling cost without improving its probability score. 
\item If $n_{r_0}S(r_0, r_i) < minSS$, then either $n_{r_i} = 0$ or $n_{r_i} = minSS - n_{r_0}S(r_0, r_i)$. This is because if $n_{r_i}$ is between $0$ and $minSS - n_{r_0}S(r_0, r_i)$, then we can decrease it to $0$, and if it is $> minSS - n_{r_0}S(r_0, r_i)$, then we can decrease it to $minSS - n_{r_0}S(r_0, r_i)$. Both these decreases would decrease sampling cost without affecting probability score. 
\squishend
Thus, there are three kinds of children $r_i$: Those with $ess \geq minSS$ but $n_{r_i} = 0$, those with $ess < minSS$ and $n_{r_i} = 0$, and those with $ess \geq minSS$ and $n_{r_i} = minSS - n_{r_0}S(r_0, r_i)$. There are $3^d$ ways to assign each child to one of these categories, and each of those potentially gives us one locally optimal solution. Consider any such locally optimal solution $e$. For $e$ let children $r_{i_1}, r_{i_2}, ... r_{i_m}$ be in the first category, $r_{i_{m+1}}, .. r_{i_M}$ be in the second category, and $r_{i_{M+1}}, .. r_{i_d}$ in the third.  Then the `probability value' of solution $e$ is given by : 
$$P(e) = \sum_{j = 1}^{i_M}p_{j}$$
and its `Sampling Cost' is
$$S(e) = \frac{minSS}{S(r_0, r_{i_m})} + \sum_{j=i_m+1}^{i_M} minSS - \frac{minSS}{S(r_0, r_{i_j})}$$
Thus, there are at most $3^d$ locally optimal solutions. $d$ is usually small (it is at most $k$, the size of the rule-list we create on every user click), even when the rule tree $U$ itself is big, and so we can enumerate all $3^d$ locally optimal solutions and find their sampling cost and probability scores.

@@

>>>>

\subsection{Dynamic Sampling for Large Tables}\label{sec:sampling}

Our greedy algorithm needs to make multiple passes over the entire table in order to find counts of rules. These passes can be rather expensive if the table is large, especially if the table does not fit in main memory. If we want exact counts for rules, we have no choice but to read the entire table.

@@

Thus, we use sampling to trade-off a small amount of accuracy for a faster response time. We now describe our sampling schemes for improving the running time of our algorithm on tables that are too large to fit in main memory.

\agp{revisit}
First in Section~\ref{sec:sample-using}, we motivate the need for sampling. Then we describe a component of our system, called the {\em SampleHandler}, which is responsible for creating and maintaining samples of the table in memory, subject to user specified memory constraints. The SampleHandler maintains multiple samples corresponding to different parts of the table, which can be used depending on which rule the user decides to expand next. Then in Section~\ref{sec:sampling_algorithms}, we describe ways to efficiently allocate memory to different samples, so as to maximize the probability that we can respond to the next user operation without accessing the hard disk. 
\papertext{In the technical report,}
\techreporttext{Finally,}
we mention some additional optimizations we can make, and
describe how we can set the minimum sample size required from the SampleHandler.

\subsubsection{Algorithms for deciding what to sample}\label{sec:sampling_algorithms}

We are given a memory capacity or budget $M$, and a minimum sample size $minSS$ (both specified by the user).
The parameter $minSS$ is the minimum number of sample tuples needed
such that we can use the sample in memory 
instead of having to resort to the entire table stored
on hard disk. 
This parameter determines how accurate our count estimates will be, and also
how quickly we can return results to the user. 
We describe how to set $minSS$ in the technical report.

We now consider the following problem: say we currently have no
samples in memory (we describe the scenario where there are already 
some samples in Section~\ref{sec:sample-handler}), and say the user is currently viewing some rules; how do we
materialize the ``best possible'' samples in memory that fit within the capacity $M$,
such that we can respond to as many user interactions using the stored samples,
without having to resort to retrieving the entire table.
That is, we want to maximize the probability that the next user interaction
can be answered using the existing samples, without reading from the hard disk. 
We leverage techniques from approximation algorithms and optimization theory, formalized
below.

\stitle{Tree of Rules.}
At any stage, we have a tree $U$ of rules displayed to the user, 
with each node of the tree corresponding to a displayed rule. 
In the rest of this section, we will refer to nodes in $U$ and rules interchangeably.
The tree is formed as follows: The root of the tree corresponds to the trivial rule. 
And suppose the user expands a node with rule $r$, resulting in rules $r_1, r_2, .. r_k$ being displayed. 
Then we add children nodes to the expanded node, corresponding to rules $r_1, r_2, .. r_k$, and so on.

Even though the rules displayed to the user can be considered a tree, 
multiple nodes of the tree may correspond to the same rule. 
For example, $(a, b)$ may be a child of both $(\star, b)$ and $(a, \star)$.
Thus, even though the structure displayed to the user is a tree, 
the set of rules displayed forms a partially ordered set (poset) using the sub-rule ordering. 
However, for the purposes of this section, we focus on the tree representation.

Internal nodes of the tree $U$ are ones that have been expanded (drilled down on), 
while the leaves are nodes that have not been expanded. 
Let $L$ be the set of leaves. 
Each leaf is something that the user can potentially expand in the next step, and thus we would like to have pre-fetched samples for rules corresponding to leaf nodes.

\stitle{Probability Distribution.} 
We assume that we have a probability distribution over leaves, which assigns a probability that each leaf will be the next one to be expanded. In the absence of additional data, we can assume a uniform probability distribution. That is, we can assume that every leaf is equally likely to be expanded next. If we have data on past user behaviour available, we can use Machine Learning on node features such as `node depth in tree', `weight of node rule' and `distance from last expanded node' to get a better probability estimate of each node being expanded next.

% When the SamplerHandler uses \textbf{Create} for a rule $r$, it needs to make a pass through the entire table, as well as potentially shrink some existing samples to free up memory. Since making a pass over the the table from the hard disk is usually a bottleneck, it can also do things like creating samples for rules other than $r$, and augmenting existing samples, in the same pass. Hence, we assume that in a \textbf{Create} phase, the SampleHandler not only creates one new sample for $r$, but also potentially creates other new samples, or resizes existing samples. For each displayed rule $r^{\prime}$, it may create a new sample for $r^{\prime}$ in the same pass through the table. Based on the current leaves and their expansion probabilities, it finds an optimal set of sizes of samples to create for each displayed rule. More specifically, for every displayed rule $r^{\prime}$, it determines an integer $n_{r^{\prime}}$, and then creates a fresh sample $s_{r^{\prime}}$ with $f_{s_{r^{\prime}}} = r^{\prime}$ and $|T_{s_{r^{\prime}}}| = n_{r^{\prime}}$ while making its pass through the table.

\stitle{Sampling Strategy.} The sampling strategy we adopt involves storing, 
for every displayed rule $r'$, a sample of $r'$ of size $n_{r'}$, i.e., 
containing $n_{r'}$ randomly chosen tuples $\in r'$.
Note that we may choose to not store a sample for some rules, in which case $n_{r'}$
will be zero for those rules. 
We need to pick the $n_{r^{\prime}}$ values so as to maximize the probability that the next user drill down can be satisfied using samples available in memory. 
As it turns out, picking a sample for some rules can help not just that rule, but
also all of its sub-rules, all the while preserving uniform randomness of samples. We formalize this notion next.

\stitle{Selectivity.} 
Let the `selectivity' of a rule be the fraction of tuples in $T$ that are covered by the rule. 
For each pair of rules $r_1, r_2 \in U$ such that $r_1$ is a sub-rule of $r_2$, we can estimate the ratio of selectivities of $r_1$ and $r_2$ using existing samples. We denote this quantity as $S(r_1, r_2)$. 
We define $S(r_1, r_2)$ to be $0$ if $r_1$ is not a sub-rule of $r_2$. 
If the same rule occurs in multiple nodes $r_1$, $r_2$ of the tree, then $S(r_1, r_2)$ is naturally $1$. 
Essentially, $S(r_1, r_2)$ denotes how much $r_1$'s sample helps $r_2$. 
Naturally, if $r_1$ is a super-rule of $r_2$, then $S$ will be $0$ since using a sample of $r_1$
for $r_2$ will lead to bias.

Then, if we have an $n_r$ sized sample with filter $r$ for each $r \in U$, the expected number of tuples covered by $r^{\prime} \in L$ is denoted as $ess(r')$ (for `effective sample size'), as below: 
\begin{equation} \label{def:ess}
ess(r^{\prime}) = \sum_{r \in U} S(r, r^{\prime})n_r
\end{equation}
Basically, $ess(r')$ captures how much of an unbiased sample of $r'$ can be retrieved using all the samples for the rules in $U$.
If $ess(r^{\prime}) \geq minSS$, then if the user expands $r^{\prime}$, we do not need to make another 
pass through the table.

We wish to maximize the probability that we can respond to the next user expansion without making another pass. We now formally define our problem below:
\begin{problem}\label{prob:sample-sizes}
Given a tree of rules $U$ with leaves $L$, a probability distribution $p$ over $L$, an integer $M$, and selectivity ratio $S(r_1, r_2)$ for each $r_1, r_2 \in U$, 
choose an integer $n_r \geq 0$ for each $r \in U$ so as to $\textrm{maximize}$ :
$$\sum_{r^{\prime} \in L} p_{r^{\prime}}I_{[ess(r^{\prime}) \geq minSS]}$$
where the $I$'s are indicator variables, subject to :
$\sum_{r \in U} n_r < M$
\end{problem}
Problem~\ref{prob:sample-sizes} is non-linear and non-convex because of the indicator variables. 
%If the tree $U$ is not too big, then it might be feasible to use an exponential algorithm to solve the problem. 
We can show that Problem~\ref{prob:sample-sizes} is {\sc NP-Hard} using a reduction from the knapsack problem. 
\begin{lemma}\label{lemma:sampling-hardness}
Problem~\ref{prob:sample-sizes} is {\sc NP-Hard}.
\end{lemma}
\begin{proof}(Sketch)
Suppose we are given an instance of the knapsack problem with $m$ objects, with the $i^{th}$ object having weight $w_i$ and value $v_i$. We are also given a weight limit $W$, and our objective is to choose a set of objects that maximizes value and has total weight $< W$. We will reduce this instance to an instance of Problem~\ref{prob:sample-sizes}.

We first scale the $w_i$s and $W$ such that all $w_i$s are $< 1$, without effectively changing the problem. For Problem~\ref{prob:sample-sizes}, we set $M$ to $(m + W) \times minSS$. Tree $U$ has $m$ special nodes $r_1, r_2, ... r_m$, and each $r_i$ has two children $r_{i,1}, r_{i,2}$. These $2m$ children are all leaves, and all leaves other than these have expansion probability $0$. 
The $S$ values are such that $\forall i \in \left\lbrace 1, 2, ..m\right\rbrace, j \in \left\lbrace 1, 2 \right\rbrace : (S(x, r_{i,j}) \neq 0 \Rightarrow x = r_{i,j} || x = r_i)$. In reality, the $S$ values cannot be exactly zero when the first argument is the trivial rule, but we can make it small enough such that any optimal solution will set $n_r = 0$ when $r$ is not a special node or its child. Therefore, each leaf $r_{i,j}$ gets tuples either from its own sample, or the sample of its parent, and from nowhere else. Thus, $ess(r_{i,j}) = n_{r_{i,j}} + n_{r_i}S(r_i, r_{i,j}) \forall 1 \leq i \leq m, j \in \left\lbrace 1,2 \right\rbrace$. In addition, $S(r_i, r_{i,1}) = 1$ (again, it cannot be exactly $1$, but can be brought arbitrarily close to $1$), and $S(r_i, r_{i,2}) = 1 - w_i$. Finally, for each i, $p_{r_{i,1}} = \frac{2}{2m+1}$ and $p_{r_{i,2}} = \frac{v_i}{(2m+1)\sum_{j=1}^{m}v_i}$. Thus, each individual $p_{r_{i,1}}$ value is higher than all $p_{r_{i,2}}$ values combined, and $M$ is high enough to cover all $r_{i,1}$. As a result, in any optimal solution, $ess(r_{i,1})$ must be equal to $minSS$, and we're left to decide which $i$'s should also have $ess(r_{i,2}) = minSS$. For all $i$, we must have either $n_{r_i} = n_{r_{i,2}} = 0 \land n_{r_{i,1}} = minSS$ (iff $ess(r_{i,2}) < minSS$) or $n_{r_i} = minSS \land n_{r_{i,1}} = 0 \land n_{r_{i,2}} = minSS (1 - S(r_i, r_{i,2})) = minSS \times w_i$ (iff $ess(r_{i,2}) = minSS$). The latter option consumes $w_i \times minSS$ extra memory and gives an extra $\frac{v_i}{(2m+1)(\sum_{j=1}^{m}v_j)}$ probability value. Thus, having $ess(r_{i,2}) = minSS$ is equivalent to picking object $i$ from the knapsack problem (as it consumes additional memory $\propto w_i$ and gives additional probability value $\propto v_i$). Moreover, the additional memory available (on top of the $m \times minSS$ required to cover all $r_{i,1}$s) is $W\times minSS$. Hence, solving Problem~\ref{prob:sample-sizes} with the above $U$, $S$, $p$ and picking the set of $i$'s for which $ess(r_{i,2}) = minSS$ gives a solution to the instance of the knapsack problem. 
\end{proof}

\stitle{Approximate DP Solution.} Even though the problem as stated is {\sc NP-Hard}, with an additional simplifying assumption, we can make the problem approximately solvable using a Dynamic Programming algorithm. The assumption is: for each $r \in L$, we assume that it will get tuples only from samples obtained for itself (filter $= r$) and its immediate parent. That is, we set $S(r_1, r_2)$ to be zero if $r_1 \neq r_2$ and $r_2$ is not a child of $r_1$. This is similar to what we had for tree $U$ in our proof of Lemma~\ref{lemma:sampling-hardness}. $ess(r)$ now becomes $ess(r^{\prime}) = n_{r^{\prime}} + n_rS(r, r^{\prime})$ where $r$ is the parent of $r^{\prime}$.

Now consider a rule in $r_0 \in U \setminus L$ along with all its children. Let $M_{r_0}$ denote the set containing $r_0$ and all its leaf children. Then, by our simplification, the number of tuples $n_{r^{\prime}}$ for any rule in $r^{\prime} \in M_{r_0}$ only affects the ess value of rules in $M_{r_0}$. This allows us to effectively split the problem into multiple subproblems,  with one subproblem per $M_{r_0}$. Thus, for each non-leaf rule $r_0$ and all its children, we compute all `locally optimal' assignments of $n_r \mid r \in M_{r_0}$. Locally optimal means that we cannot get a higher value of `probability value' $\sum_{r \in M_{r_0}} p_rI_{ess(r) \geq minSS}$ for the same `sampling cost' $\sum_{r\in M_{r_0}} n_r$. Then, we can use dynamic programming to combine the locally optimal solutions of different $M_{r_0}$s. We describe both these steps in detail below:

Let $r_0 \in U \setminus L$. Let $d$ be the number of leaf children of $r_0$. Let the children be $r_1, r_2, ... r_d$. For any child $r_i$, $n_{r_i}$ only contributes to its own $ess$, whereas $n_{r_0}$ contributes to the $ess$ of all children $r_1, ... r_d$. Given a value of $n_{r_0}$, in a locally optimal solution, each child $r_i$ must satisfy:
\squishlist
\item If $n_{r_0}S(r_0, r_i) \geq minSS$, then $n_{r_i} = 0$ because otherwise, decreasing $n_{r_i}$ to $0$ would lower its sampling cost without improving its probability score. 
\item If $n_{r_0}S(r_0, r_i) < minSS$, then either $n_{r_i} = 0$ or $n_{r_i} = minSS - n_{r_0}S(r_0, r_i)$. This is because if $n_{r_i}$ is between $0$ and $minSS - n_{r_0}S(r_0, r_i)$, then we can decrease it to $0$, and if it is $> minSS - n_{r_0}S(r_0, r_i)$, then we can decrease it to $minSS - n_{r_0}S(r_0, r_i)$. Both these decreases would decrease sampling cost without affecting probability score. 
\squishend
Thus, there are three kinds of children $r_i$: Those with $ess \geq minSS$ but $n_{r_i} = 0$, those with $ess < minSS$ and $n_{r_i} = 0$, and those with $ess \geq minSS$ and $n_{r_i} = minSS - n_{r_0}S(r_0, r_i)$. There are $3^d$ ways to assign each child to one of these categories, and each of those potentially gives us one locally optimal solution. Consider any such locally optimal solution $e$. For $e$ let children $r_{i_1}, r_{i_2}, ... r_{i_m}$ be in the first category, $r_{i_{m+1}}, .. r_{i_M}$ be in the second category, and $r_{i_{M+1}}, .. r_{i_d}$ in the third.  Then the `probability value' of solution $e$ is given by : 
$P(e) = \sum_{j = 1}^{i_M}p_{j}$,
and its `Sampling Cost' is
$$S(e) = \frac{minSS}{S(r_0, r_{i_m})} + \sum_{j=i_m+1}^{i_M} minSS - \frac{minSS}{S(r_0, r_{i_j})}$$
Thus, there are at most $3^d$ locally optimal solutions; $d$ is usually small (it is at most $k$, the size of the rule-list we create on every user click), even when the rule tree $U$ itself is big, and so we can enumerate all $3^d$ locally optimal solutions and find their sampling cost and probability scores.

@@

\subsubsection{Design of the SampleHandler}\label{sec:sample-handler}
We now describe the design of the {\em SampleHandler}, which given
a certain memory capacity $M$, and a minimum sample size $minSS$,
creates, maintains, retrieves, and removes samples, all in
response to user interactions on the table.
It uses the algorithms from Section~\ref{sec:sampling_algorithms} 
to decide which samples to create, as we will see below.

% Example from situation in a previous table. And good and bad samples to keep there.
At all points, the SampleHandler maintains a set of samples in memory. For instance, it may keep a sample of tuples used to expand the first (trivial) rule, and another sample used to expand the rule last clicked on by the user. 
Each sample $s$ is represented as a triple: (a) A `filter' rule $f_s$, (b) a scaling factor $N_s$ and (c) a set $T_s$ of tuples from the table. The set $T_s$ consists of a $\frac{1}{N_s}$ uniformly sampled fraction of tuples covered by $f_s$. The scaling factor $N_s$ is used to translate the count of a rule on the sample into an estimate of the count over the entire table. The sum of $|T_s|$ over all samples $s$ is not allowed to exceed capacity $M$ at any point. 
% TODO: Give example of what samples we'd keep.
